Revealing transient strain in geodetic data with Gaussian process regression

### 2017-09-28 RevisedSubmission/hines_hetland_2017_gji.tex
did more shit

The posterior transient displacements are spatially and temporally continuous, and we can use eqs. (\ref{eq:PosteriorMean3}) and (\ref{eq:PosteriorCov3}) to evaluate $\hat{u}_i$ at any $p$. Furthermore, $\hat{u}_i$ is spatially and temporally differentiable, allowing us to formulate $\dot\varepsilon$ at any $p$ that we may be interested in. The components of $\dot\varepsilon$ can be written as 
\begin{equation}\label{eq:StrainRate}
\dot\varepsilon_{ij}(p) = \frac{1}{2} \frac{\partial}{\partial t} \left(
                                        \frac{\partial \hat{u}_i(p)}{\partial x_j} +  
                                        \frac{\partial \hat{u}_j(p)}{\partial x_i}\right).
\end{equation}
Since eq. (\ref{eq:StrainRate}) is a linear operation on the Gaussian processes $u_i$ and $u_j$, we know that $\dot\varepsilon_{ij}$ is also a Gaussian process. The mean and covariance functions for $\dot{\varepsilon}_{ij}$ are given by
\begin{equation}\label{eq:StrainMean}
\mu_{\dot\varepsilon_{ij}}(p) = \frac{1}{2}\frac{\partial}{\partial t}\left(
                                  \frac{\partial \mu_{\hat{u}_i}(p)}{\partial x_j} + 
                                  \frac{\partial \mu_{\hat{u}_j}(p)}{\partial x_i} \right)
\end{equation} 
and
\begin{equation}\label{eq:StrainCov}
C_{\dot\varepsilon_{ij}}(p,p') = \frac{1}{4} \frac{\partial^2}{\partial t \, \partial t'}\left(
                                   \frac{\partial^2 C_{\hat{u}_i}(p,p')}{\partial x_j \, \partial x'_j} + 
                                   \frac{\partial^2 C_{\hat{u}_j}(p,p')}{\partial x_i \, \partial x'_i} \right),
\end{equation} 
respectively.

Our motivation for estimating transient strain rates is, in part, to detect geophysical phenomena. As we will see, geophysical signal can be easily identified by visually inspecting the solution for transient strain rates. However, if we want to detect geophysical phenomena automatically, then we need to define a detection criterion. We use a signal-to-noise ratio, SNR, that is based on the Frobenius norm of $\dot\varepsilon$, $||\dot\varepsilon||_\mathrm{F} = \left(\dot\varepsilon_{\mathrm{ee}}^2 + \dot\varepsilon_{\mathrm{nn}}^2 + 2\dot\varepsilon_{\mathrm{en}}^2\right)^{\frac{1}{2}}$, for our detection criterion. In the geodetic literature, $||\dot{\varepsilon}||_\mathrm{F}$ is often used as a metric for the strain rate ``magnitude", and it is sometimes referred to as the second invariant of strain rate. Noting that $||\dot\varepsilon||_\mathrm{F}$ is a random variable, we take SNR to be the ratio of the estimated mean and standard deviation of $||\dot\varepsilon||_\mathrm{F}$. An estimate of the mean is found by evaluating $||\dot\varepsilon||_\mathrm{F}$ at the mean of $\dot\varepsilon$, 
\begin{align}
\mu_{||\dot\varepsilon||_\mathrm{F}} &\approx \left. ||\dot\varepsilon||_F \right|_{\dot\varepsilon = \mu_{\dot\varepsilon}} \\
                                     &= \left(\mu_{\dot\varepsilon_\mathrm{nn}}^2 + 
                                              \mu_{\dot\varepsilon_\mathrm{ee}}^2 + 
                                              2\mu_{\dot\varepsilon_\mathrm{en}}^2
                                     \right)^\frac{1}{2},
\end{align}
and we use nonlinear uncertainty propagation to estimate the standard deviation of $||\dot\varepsilon||_\mathrm{F}$,
\begin{equation}
\sigma_{||\dot\varepsilon||_\mathrm{F}} \approx
\left(\left(\left. \frac{\partial ||\dot\varepsilon||_\mathrm{F}}{\partial \dot\varepsilon_\mathrm{nn}}\right|_{\dot\varepsilon = \mu_{\dot\varepsilon}} \right)^2 \sigma^2_{\varepsilon_\mathrm{nn}} +
\left(\left. \frac{\partial ||\dot\varepsilon||_\mathrm{F}}{\partial \dot\varepsilon_\mathrm{ee}}\right|_{\dot\varepsilon = \mu_{\dot\varepsilon}} \right)^2 \sigma^2_{\varepsilon_\mathrm{ee}} +
\left(\left. \frac{\partial ||\dot\varepsilon||_\mathrm{F}}{\partial \dot\varepsilon_\mathrm{en}}\right|_{\dot\varepsilon = \mu_{\dot\varepsilon}} \right)^2 \sigma^2_{\varepsilon_\mathrm{en}}
\right)^\frac{1}{2},
\end{equation}
where $\sigma^2_{\varepsilon_{ij}}(p) = C_{\varepsilon_{ij}}(p,p)$. 
 by evaluating it at $\dot\varepsilon_{ij} = \mu_{\dot\varepsilon_{ij}}$. point $p$ is approximately $(\mu_{\dot\varepsilon_\mathrm{nn}}(p) + \mu_{\dot\varepsilon_\mathrm{ee}}(p) + 2\mu_{\dot\varepsilon_\mathrm{en}}(p))$. Using nonlinear uncertainty propagation, the standard deviation of $||\dot\varepsilon||$ is approximated as

can be found with nonlinear uncertainty propagation,

 Using nonlinear uncertainty propagation, we find SNR to be  
\begin{equation}\label{eq:SNR}
\mathrm{SNR}(p) = \frac{\mu_{\dot\varepsilon_\mathrm{nn}}(p)^2 +
                        \mu_{\dot\varepsilon_\mathrm{ee}}(p)^2 +
                        2\mu_{\dot\varepsilon_\mathrm{en}}(p)^2}
                       {\big(C_{\dot\varepsilon_\mathrm{nn}}(p,p)\mu_{\dot\varepsilon_\mathrm{nn}}(p)^2 + 
                              C_{\dot\varepsilon_\mathrm{ee}}(p,p)\mu_{\dot\varepsilon_\mathrm{ee}}(p)^2 + 
                              4C_{\dot\varepsilon_\mathrm{en}}(p,p)\mu_{\dot\varepsilon_\mathrm{en}}(p)^2
                        \big)^{\frac{1}{2}}}
,
\end{equation}
For simplicity, we have ignored covariances between the transient strain rate components in eq. (\ref{eq:SNR}), even though they are non-zero.

\section{Outlier detection}\label{sec:Outlier}
In our formulation for estimating transient strain rates, we have assumed that noise in the data vector is normally distributed. This is not an appropriate assumption for GNSS data, which often have more outliers than would be predicted for normally distributed noise. It follows that proposed methods for analyzing GNSS data should be robust against outliers \citep[e.g.,][]{Blewitt2016}. In order to make our estimates of transient strain more robust, we automatically identify and remove outliers in the GNSS data as a pre-processing step.

>>>>

The posterior transient displacements are spatially and temporally continuous, and we can use eqs. (\ref{eq:PosteriorMean3}) and (\ref{eq:PosteriorCov3}) to evaluate $\hat{u}_i$ at any $p$. Furthermore, $\hat{u}_i$ is spatially and temporally differentiable, allowing us to formulate $\dot\varepsilon$ at any $p$ that we may be interested in. The components of $\dot\varepsilon$ can be written as 
\begin{equation}\label{eq:StrainRate}
\dot\varepsilon_{ij}(p) = \frac{1}{2} \frac{\partial}{\partial t} \left(
                                        \frac{\partial \hat{u}_i(p)}{\partial x_j} +  
                                        \frac{\partial \hat{u}_j(p)}{\partial x_i}\right).
Since eq. (\ref{eq:StrainRate}) is a linear operation on the Gaussian processes $\hat{u}_i$ and $\hat{u}_j$, we know that $\dot\varepsilon_{ij}$ is also a Gaussian process. The mean and covariance functions for $\dot{\varepsilon}_{ij}$ are given by
\begin{equation}\label{eq:StrainMean}
\mu_{\dot\varepsilon_{ij}}(p) = \frac{1}{2}\frac{\partial}{\partial t}\left(
                                  \frac{\partial \mu_{\hat{u}_i}(p)}{\partial x_j} + 
                                  \frac{\partial \mu_{\hat{u}_j}(p)}{\partial x_i} \right)
\begin{equation}\label{eq:StrainCov}
C_{\dot\varepsilon_{ij}}(p,p') = \frac{1}{4} \frac{\partial^2}{\partial t \, \partial t'}\left(
                                   \frac{\partial^2 C_{\hat{u}_i}(p,p')}{\partial x_j \, \partial x'_j} + 
                                   \frac{\partial^2 C_{\hat{u}_j}(p,p')}{\partial x_i \, \partial x'_i} \right),
respectively.
\subsection{Transient detection criterion}\label{sec:TransientDetection}
Our motivation for estimating transient strain rates is, in part, to detect geophysical phenomena. As we will see, geophysical signal can be easily identified by visually inspecting the solution for $\dot\varepsilon$ from eqs. (\ref{eq:StrainMean}) and (\ref{eq:StrainCov}). However, if we want to detect geophysical phenomena automatically, then we need to define a detection criterion. We use a signal-to-noise ratio, SNR, that is based on the Frobenius norm of $\dot\varepsilon$, $||\dot\varepsilon||_\mathrm{F} = \left(\dot\varepsilon_{\mathrm{ee}}^2 + \dot\varepsilon_{\mathrm{nn}}^2 + 2\dot\varepsilon_{\mathrm{en}}^2\right)^{\frac{1}{2}}$, for our detection criterion. In the geodetic literature, $||\dot{\varepsilon}||_\mathrm{F}$ is often used as a metric for the strain rate ``magnitude", and it is sometimes referred to as the second invariant of strain rate. Noting that $||\dot\varepsilon||_\mathrm{F}$ is a random variable, we take SNR to be the ratio of the estimated mean and standard deviation of $||\dot\varepsilon||_\mathrm{F}$. An estimate of the mean is found by evaluating $||\dot\varepsilon||_\mathrm{F}$ at the mean of $\dot\varepsilon$, 
\begin{align}
\mu_{||\dot\varepsilon||_\mathrm{F}} &\approx ||\dot\varepsilon||_F \big|_{\dot\varepsilon = \mu_{\dot\varepsilon}} \\
                                     &= \left(\mu_{\dot\varepsilon_\mathrm{ee}}^2 + 
                                              \mu_{\dot\varepsilon_\mathrm{nn}}^2 + 
                                              2\mu_{\dot\varepsilon_\mathrm{en}}^2
                                     \right)^\frac{1}{2},
\end{align}
and we use nonlinear uncertainty propagation to estimate the standard deviation,
\begin{equation}
\sigma_{||\dot\varepsilon||_\mathrm{F}} \approx
\left(\left(\left. \frac{\partial ||\dot\varepsilon||_\mathrm{F}}{\partial \dot\varepsilon_\mathrm{ee}}\right|_{\dot\varepsilon = \mu_{\dot\varepsilon}} \right)^2 \sigma^2_{\dot\varepsilon_\mathrm{ee}} +
\left(\left. \frac{\partial ||\dot\varepsilon||_\mathrm{F}}{\partial \dot\varepsilon_\mathrm{nn}}\right|_{\dot\varepsilon = \mu_{\dot\varepsilon}} \right)^2 \sigma^2_{\dot\varepsilon_\mathrm{nn}} +
\left(\left. \frac{\partial ||\dot\varepsilon||_\mathrm{F}}{\partial \dot\varepsilon_\mathrm{en}}\right|_{\dot\varepsilon = \mu_{\dot\varepsilon}} \right)^2 \sigma^2_{\dot\varepsilon_\mathrm{en}}
\right)^\frac{1}{2},
where $\sigma^2_{\dot\varepsilon_{ij}}(p) = C_{\dot\varepsilon_{ij}}(p,p)$. After some calculations, we find SNR to be
\begin{align}\label{eq:SNR}
\mathrm{SNR}(p) &= \frac{\mu_{||\dot\varepsilon||_\mathrm{F}}(p)}{\sigma_{||\dot\varepsilon||_\mathrm{F}}(p)} \\
                &= \frac{\mu_{\dot\varepsilon_\mathrm{ee}}(p)^2 +
                        \mu_{\dot\varepsilon_\mathrm{nn}}(p)^2 +
                        2\mu_{\dot\varepsilon_\mathrm{en}}(p)^2}
                       {\big(\sigma^2_{\dot\varepsilon_\mathrm{ee}}(p)\mu_{\dot\varepsilon_\mathrm{ee}}(p)^2 + 
                             \sigma^2_{\dot\varepsilon_\mathrm{nn}}(p)\mu_{\dot\varepsilon_\mathrm{nn}}(p)^2 + 
                              4\sigma^2_{\dot\varepsilon_\mathrm{en}}(p)\mu_{\dot\varepsilon_\mathrm{en}}(p)^2
                        \big)^{\frac{1}{2}}}
.
\end{align}
We explicitly show that SNR is a function of $p$ to emphasize that it identifies the position and time of anomalous deformation. We can reasonably suspect that some transient geophysical phenomena is occurring wherever and whenever SNR is larger than ${\sim}3$.

\subsection{Outlier detection}\label{sec:Outlier}
In our formulation for estimating transient strain rates, we have assumed that noise in the data vector is normally distributed. This is not an appropriate assumption for GNSS data, which often have more outliers than would be predicted for normally distributed noise. It follows that proposed methods for analyzing GNSS data should be robust against outliers \citep[e.g.,][]{Blewitt2016}. In order to make our estimates of transient strain more robust, we automatically identify and remove outliers in the GNSS data as a pre-processing step.
